---
title: "Take-Home_Ex1"
author: "Roger"
editor: visual
---

# 1.1 Overview

Water is an important resource to mankind. Clean and accessible water is critical to human health. It provides a healthy environment, a sustainable economy, reduces poverty and ensures peace and security. Yet over 40% of the global population does not have access to sufficient clean water. By 2025, 1.8 billion people will be living in countries or regions with absolute water scarcity, according to UN-Water. The lack of water poses a major threat to several sectors, including food security. Agriculture uses about 70% of the world's accessible freshwater.

Developing countries are most affected by water shortages and poor water quality. Up to 80% of illnesses in the developing world are linked to inadequate water and sanitation. Despite technological advancement, providing clean water to the rural community is still a major development issues in many countries globally, especially countries in the Africa continent.

To address the issue of providing clean and sustainable water supply to the rural community, a global [Water Point Data Exchange (WPdx)](https://www.waterpointdata.org/about/) project has been initiated. The main aim of this initiative is to collect water point related data from rural areas at the water point or small water scheme level and share the data via WPdx Data Repository, a cloud-based data library. What is so special of this project is that data are collected based on [WPDx Data Standard](https://www.waterpointdata.org/wp-content/uploads/2021/04/WPDx_Data_Standard.pdf).

## 1.1.1 Objectives

Geospatial analytics hold tremendous potential to address complex problems facing society. In this study, appropriate global and local measures of spatial Association techniques will be applied to reveal the spatial patterns of Not Functional water points. For the purpose of this study, Nigeria will be used as the study country.

In the code chunk below, p_load(0 of pacman package is used to load the following R packages into R environment:

-   sf

-   spdep

-   tmap

-   tidyverse

-   funModeling (used for rapid Exploratory Data Analysis)

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, funModeling)
```

# 1.2 Data Acquisition

## 1.2.1 Importing the geospatial data

For this study, two geospatial data will be used:

### 1.2.1.1 Importing water point geospatial data

-   The water point geospatial data will be downloaded from [WPdx Global Data Repositories](https://www.waterpointdata.org/access-data/), specifically the WPdx+ data set will be used. In the code chunk below, we will import the shapefile as simple features data table into R environment using st_read() of sf package. filter() of dplyr will be used to extract water point records of Nigeria.

```{r}
#| eval: false

wp <- st_read(dsn = "data",
              layer= "geo_export",
              crs = 4326) %>%
  filter(clean_coun == "Nigeria")
```

Next, write_rds() of readr package is used to save the extracted sf data table (i.e. wp) into an output file in rds data format, and saved in the data sub-folder.

```{r}
#| eval: false

write_rds(wp, "data/wp_nga.rds")
```

### 1.2.1.2 Importing Nigeria LGA boundary geospatial data

-   The Nigeria Level-2 Administrative Boundary (also known as Local Government Area) polygon features GIS data will be the second data set used in this study. The data is downloaded from [geoBoundaries](https://www.geoboundaries.org/). In the code chunk below, we will import the Nigeria LGA boundary shapefile data as a simple features data table into R environment using st_read() of sf package.

```{r}
#| eval: false

nga <- st_read(dsn = "data", 
                 layer = "nga_polnda_adm2_1m_salb",
                 crs = 4326) 
```

## 1.2.2 Data Wrangling

### 1.2.2.1 Recoding NA values into string

In the code chunk below, replace_na() of tidyr package is used to recode all the NA values in *status_cle* field into "Unknown".

```{r}
#| eval: false

wp_nga <- read_rds("data/wp_nga.rds") %>%
  mutate(status_cle = replace_na(status_cle, "Unknown"))
```

### 1.2.2.2 EDA

In the code chunks below, freq() of funModeling package is used to display the distribution of *status_cle* field in *wp_nga*.

```{r}
#| eval: false

freq(data = wp_nga,
     input = "status_cle")
```

### 1.2.2.3 Extracting functional water point

In the code chunk below, filter() of dplyr is used to select functional water points.

```{r}
#| eval: false

wpt_functional <- wp_nga %>%
  filter(status_cle %in%
           c("Functional",
             "Functional but not in use",
             "Functional but needs repair"))
```

```{r}
#| eval: false

freq(data= wpt_functional,
     input = "status_cle")
```

### 1.2.2.4 Extracting non-functional water point

In the code chunk below, filter() of dplyr is used to select non-functional water points.

```{r}
#| eval: false

wpt_nonfunctional <- wp_nga %>%
  filter(status_cle %in%
           c("Abandoned/Decommissioned",
             "Abandoned",
             "Non-Functional",
             "Non functional due to dry season",
             "Non-Functional due to dry season"))
```

```{r}
#| eval: false

freq(data = wpt_nonfunctional,
     input = "status_cle")
```

### 1.2.2.5 Extracting water point with Unknown Class

In the code chunk below, filter() of dplyr is used to select water points with unknown status.

```{r}
#| eval: false

wpt_unknown <- wp_nga %>%
  filter(status_cle == "Unknown")
```

### 1.2.2.6 Performing Point-in-Polygon Count

In the code chunk below, st_intersects() of sf is used to identify water points located in each Nigerian province; *lengths()* of Base R is used to calculate numbers of water points that fall inside each province, and mutate() of dplyr package is used to derive four fields, namely: (a) total wpt, (b) wpt functional, (c) wpt non-functional, (d) wpt unknown.

```{r}
#| eval: false

nga_wp <- nga %>% 
  mutate(`total wpt` = lengths(
    st_intersects(nga, wp_nga))) %>%
  mutate(`wpt functional` = lengths(
    st_intersects(nga, wpt_functional))) %>%
  mutate(`wpt non-functional` = lengths(
    st_intersects(nga, wpt_nonfunctional))) %>%
  mutate(`wpt unknown` = lengths(
    st_intersects(nga, wpt_unknown)))
```

### 1.2.2.7 Saving the Analytical Data Table

In the code chunk below, mutate() of dplyr package is used to drive two fields, namely pct_functional and pct non-functional. To keep the file size small, select() of dplyr is used to retain only fields 3,4,9,10,18,19,20,21,22 and 23.

```{r}
#| eval: false

nga_wp <- nga_wp %>% 
  mutate(`pct_functional` = `wpt functional` / `total wpt`) %>%
  mutate(`pct_non-functional` = `wpt non-functional`/ `total wpt`) #%>%
#  select(3:4, 9:10, 18:23)
```

Next, write_rds() of readr package is used to save the extracted sf data table into an output file in rds data format, and saved in the data sub-folder.

```{r}
#| eval: false

write_rds(nga_wp, "data/nga_wp.rds")
```

```{r}
nga_wp <- read_rds("data/nga_wp.rds")
total <- qtm(nga_wp, "total wpt")
wp_functional <- qtm(nga_wp, "wpt functional")
wp_nonfunctional <- qtm(nga_wp, "wpt non-functional")
unknown <- qtm(nga_wp, "wpt unknown")

tmap_arrange(total, wp_functional, wp_nonfunctional, unknown, asp=1, ncol=2)
```
